---
title: "Spectra plot: base R plot vs ggplot2"
format: html
---

# Base R plot vs ggplot2

```{r, echo=FALSE}
library(data.table)
library(profmem)
lib_base <- "test_env_base/library"
lib_ggplot <- "test_env_ggplot/library"
summarise_mb <- function(mb, label) {
    df <- as.data.table(mb)[, .(median = median(time)/10^9), by = "expr"]
    df$expr <- as.character(df$expr)
    df$engine <- label
    df$expr <- factor(df$expr, levels = sort(df$expr))
    df
}

```

In this document we will compare the Spectra plotting functions based on:

- Base R plot
- ggplot2

## Installation

First we compare how the ggplot/ggiraph dependencies affect the installation of
Spectra. To do so, we will do a clean new installation of Spectra in two
different folders.
The installation with ggplot2 require many more dependencies.

```{r, message = FALSE}
## Prepare the folders
dir.create(lib_base, recursive = TRUE, showWarnings = FALSE)
dir.create(lib_ggplot, recursive = TRUE, showWarnings = FALSE)

benchmark_install <- function(repo, lib_path, label = "") {
    ## Clean downloaded package.
    pak::pak_cleanup(force = TRUE)

    BiocManager::install(version = "devel", update = FALSE, ask = FALSE)
    options(repos = BiocManager::repositories())

    t0 <- Sys.time()
    result <- pak::pkg_install(
        repo,
        lib = lib_path,
        ask = FALSE,
        dependencies = TRUE,
        upgrade = TRUE
    )
    t1 <- Sys.time()

    list(
        label = label,
        time_sec = as.numeric(difftime(t1, t0, units = "secs")),
        installed_pkgs = list.files(lib_path),
        n_pkgs = length(list.files(lib_path))
    )
}

res_base_install <- suppressMessages(benchmark_install("rformassspectrometry/Spectra",
                                        lib_base, label = "Base"))

res_ggplot_install <- suppressMessages(benchmark_install("gabrieletome/Spectra",
                                        lib_ggplot, label = "ggPlot2"))

comparison <- data.frame(
    label = c(res_base_install$label, res_ggplot_install$label),
    time_seconds = c(res_base_install$time_sec, res_ggplot_install$time_sec),
    n_packages_installed = c(res_base_install$n_pkgs, res_ggplot_install$n_pkgs))
print(comparison)
```

## Run the plotting benchmarks

The two Spectra installs (`r lib_base` and `r lib_ggplot`) cannot both be
loaded in this same R session.
Instead, each half of the benchmark is a **separate `.qmd` file**
(`_bench_base.qmd`, `_bench_ggplot.qmd`), and we render each of them with its
own `quarto render` call. In this way we can compare the correctly the two
function.

```{r, results='hide'}
system2("quarto", c("render", "_bench_base.qmd", "--to", "html"))
system2("quarto", c("render", "_bench_ggplot.qmd", "--to", "html"))
```

## Compare the results: base plot VS ggplot

Each child render saved its `microbenchmark` result objects to disk
(`benchmark_plot.Rdata`, `benchmark_ggplot.Rdata`). We load both back into
this session to build the final comparison.

For the comparison we will use the same dataset used in the Spectra vignette.
The data are loaded in memory using `MsBackendMemory()`, in this way we will
check only the difference between base R plot and ggplot. Later in the vignette
a smaller test based on `MsBackendMzR()` will be conducted.

```{r}
load("benchmark_plot.Rdata")
load("benchmark_ggplot.Rdata")
```

- Each function runs 10 times to obtain a solid results.

### Compare load time

```{r}
plot_load <- rbind(
    summarise_mb(res_plot_load, "base R"),
    summarise_mb(res_ggplot_load, "ggplot2")
)
plot_load
```

### `plotSpectra()`/`ggplotSpectra()` with `MsBackendMemory()`

We start with the function `plotSpectra()` with 3 runs:
- Plot a single spectra.
- Plot 3 spectra.
- Plot 11 spectra.
- Plot 101 spectra.

With 101 spectra gpglot2 generate the plot instead the base R function fail
since exced the size limit.

```{r}
plot_summary <- rbind(
    summarise_mb(res_plot_t1, "base R"),
    summarise_mb(res_ggplot_t1, "ggplot2")
)
plot_summary
```

```{r}
library(ggplot2)

ggplot(plot_summary, aes(x = expr, y = median, fill = engine)) +
    geom_col(position = "dodge") +
    labs(title = "plotSpectra() vs ggplotSpectra(): median time (s)",
         x = NULL, y = "seconds") +
    coord_flip()
```

### `plotSpectraMirror()`/`ggplotSpectraMirror()` with `MsBackendMemory()`

In this section we test the mirroring plot with 2 random spectra.

```{r}
mirror_summary <- rbind(
    summarise_mb(res_plot_mirror_t1, "base R"),
    summarise_mb(res_ggplot_mirror_t1, "ggplot2")
)
mirror_summary
```

```{r}

ggplot(mirror_summary, aes(x = expr, y = median, fill = engine)) +
    geom_col(position = "dodge") +
    labs(title = "plotSpectraMirror() vs ggplotSpectraMirror(): median time (s)",
         x = NULL, y = "seconds") +
    coord_flip()
```

### `plotSpectraOverlay()`/`ggplotSpectraOverlay()` with `MsBackendMemory()`

Now we test the overlay plot. We perform 5 tests with different subset:
- Overlay with 1 spectra.
- Overlay with 3 spectra.
- Overlay with 11 spectra.
- Overlay with 101 spectra.
- Overlay with all spectra.

```{r}
overlay_summary <- rbind(
    summarise_mb(res_plot_overlay_t1, "base R"),
    summarise_mb(res_ggplot_overlay_t1, "ggplot2")
)
overlay_summary
```

```{r}
ggplot(overlay_summary[median < 5], aes(x = expr, y = median, fill = engine)) +
    geom_col(position = "dodge") +
    labs(title = "plotSpectraOverlay() vs ggplotSpectraOverlay(): median time (s)",
         x = NULL, y = "seconds") +
    coord_flip()
```

```{r}
ggplot(overlay_summary[median > 5], aes(x = expr, y = median, fill = engine)) +
    geom_col(position = "dodge") +
    labs(title = "plotSpectraOverlay() vs ggplotSpectraOverlay(): median time (s)",
         x = NULL, y = "seconds") +
    coord_flip()
```

### `plotSpectra()`/`ggplotSpectra()` with `MsBackendMzR()`

We start with the function `plotSpectra()` with 3 runs:
- Plot a single spectra.
- Plot 3 spectra.
- Plot 11 spectra.

```{r}
plot_summary_mzr <- rbind(
    summarise_mb(res_plot_t1_mzr, "base R"),
    summarise_mb(res_ggplot_t1_mzr, "ggplot2")
)
plot_summary_mzr
```

```{r}
ggplot(plot_summary_mzr, aes(x = expr, y = median, fill = engine)) +
    geom_col(position = "dodge") +
    labs(title = "plotSpectra() vs ggplotSpectra() with MsBackendMzr(): median time (s)",
         x = NULL, y = "seconds") +
    coord_flip()
```

### `plotSpectraMirror()`/`ggplotSpectraMirror()` with `MsBackendMzR()`

In this section we test the mirroring plot with 2 random spectra.

```{r}
mirror_summary_mzr <- rbind(
    summarise_mb(res_plot_mirror_t1_mzr, "base R"),
    summarise_mb(res_ggplot_mirror_t1_mzr, "ggplot2")
)
mirror_summary_mzr
```

```{r}

ggplot(mirror_summary_mzr, aes(x = expr, y = median, fill = engine)) +
    geom_col(position = "dodge") +
    labs(title = "plotSpectraMirror() vs ggplotSpectraMirror() with MsBackendMzr(): median time (s)",
         x = NULL, y = "seconds") +
    coord_flip()
```

### `plotSpectraOverlay()`/`ggplotSpectraOverlay()` with `MsBackendMzR()`

Now we test the overlay plot. We perform 5 tests with different subset:
- Overlay with 1 spectra.
- Overlay with 3 spectra.
- Overlay with 11 spectra.

```{r}
overlay_summary_mzr <- rbind(
    summarise_mb(res_plot_overlay_t1_mzr, "base R"),
    summarise_mb(res_ggplot_overlay_t1_mzr, "ggplot2")
)
overlay_summary_mzr
```

```{r}
ggplot(overlay_summary_mzr, aes(x = expr, y = median, fill = engine)) +
    geom_col(position = "dodge") +
    labs(title = "plotSpectraOverlay() vs ggplotSpectraOverlay() with MsBackendMzr(): median time (s)",
         x = NULL, y = "seconds") +
    coord_flip()
```

## SessionInfo

```{r}
sessionInfo()
```
